---
title: "R Notebook for JA timecourse 2022"
output: html_notebook
---

https://bioconductor.org/packages/release/bioc/manuals/edgeR/man/edgeR.pdf

https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf

```
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("edgeR")
BiocManager::install("topGO", force = TRUE) #exclude update of vctrs and xfun
```


```{r}
library(tidyverse)
library(edgeR)
library(pheatmap)
library(viridis)
library(gridExtra)
library(seqinr)
library(topGO)
library(GO.db)
library(dtwclust)
library(UpSetR)
```

```{r}
sessionInfo()
```

Output:
```
R version 4.3.1 (2023-06-16 ucrt)
Platform: x86_64-w64-mingw32/x64 (64-bit)
Running under: Windows 10 x64 (build 19045)

Matrix products: default


Random number generation:
 RNG:     L'Ecuyer-CMRG 
 Normal:  Inversion 
 Sample:  Rejection 
 
locale:
[1] LC_COLLATE=English_United States.utf8 
[2] LC_CTYPE=English_United States.utf8   
[3] LC_MONETARY=English_United States.utf8
[4] LC_NUMERIC=C                          
[5] LC_TIME=English_United States.utf8    

time zone: America/Halifax
tzcode source: internal

attached base packages:
[1] stats4    stats     graphics  grDevices
[5] utils     datasets  methods   base     

other attached packages:
 [1] UpSetR_1.4.0         seqinr_4.2-30       
 [3] gridExtra_2.3        viridis_0.6.4       
 [5] viridisLite_0.4.2    pheatmap_1.0.12     
 [7] lubridate_1.9.3      forcats_1.0.0       
 [9] stringr_1.5.0        dplyr_1.1.3         
[11] purrr_1.0.2          readr_2.1.4         
[13] tidyr_1.3.0          tibble_3.2.1        
[15] ggplot2_3.4.4        tidyverse_2.0.0     
[17] topGO_2.52.0         SparseM_1.81        
[19] GO.db_3.17.0         AnnotationDbi_1.62.2
[21] IRanges_2.34.1       S4Vectors_0.38.2    
[23] Biobase_2.60.0       graph_1.78.0        
[25] BiocGenerics_0.46.0  edgeR_3.42.4        
[27] limma_3.56.2         dtwclust_5.5.12     
[29] dtw_1.23-1           proxy_0.4-27        

loaded via a namespace (and not attached):
 [1] DBI_1.1.3              
 [2] bitops_1.0-7           
 [3] rlang_1.1.1            
 [4] magrittr_2.0.3         
 [5] ade4_1.7-22            
 [6] clue_0.3-65            
 [7] flexclust_1.4-1        
 [8] matrixStats_1.0.0      
 [9] compiler_4.3.1         
[10] RSQLite_2.3.1          
[11] png_0.1-8              
[12] vctrs_0.6.4            
[13] reshape2_1.4.4         
[14] pkgconfig_2.0.3        
[15] crayon_1.5.2           
[16] fastmap_1.1.1          
[17] XVector_0.40.0         
[18] ellipsis_0.3.2         
[19] utf8_1.2.3             
[20] promises_1.2.1         
[21] rmarkdown_2.25         
[22] tzdb_0.4.0             
[23] bit_4.0.5              
[24] xfun_0.40              
[25] modeltools_0.2-23      
[26] zlibbioc_1.46.0        
[27] cachem_1.0.8           
[28] GenomeInfoDb_1.36.4    
[29] blob_1.2.4             
[30] later_1.3.1            
[31] parallel_4.3.1         
[32] cluster_2.1.4          
[33] R6_2.5.1               
[34] stringi_1.7.12         
[35] RColorBrewer_1.1-3     
[36] Rcpp_1.0.11            
[37] iterators_1.0.14       
[38] knitr_1.44             
[39] timechange_0.2.0       
[40] httpuv_1.6.11          
[41] Matrix_1.5-4.1         
[42] tidyselect_1.2.0       
[43] rstudioapi_0.15.0      
[44] yaml_2.3.7             
[45] codetools_0.2-19       
[46] lattice_0.21-8         
[47] plyr_1.8.9             
[48] shiny_1.7.5.1          
[49] withr_2.5.1            
[50] KEGGREST_1.40.1        
[51] evaluate_0.22          
[52] RcppParallel_5.1.7     
[53] Biostrings_2.68.1      
[54] pillar_1.9.0           
[55] foreach_1.5.2          
[56] shinyjs_2.1.0          
[57] generics_0.1.3         
[58] RCurl_1.98-1.12        
[59] hms_1.1.3              
[60] munsell_0.5.0          
[61] scales_1.2.1           
[62] xtable_1.8-4           
[63] class_7.3-22           
[64] glue_1.6.2             
[65] tools_4.3.1            
[66] RSpectra_0.16-1        
[67] locfit_1.5-9.8         
[68] grid_4.3.1             
[69] colorspace_2.1-0       
[70] GenomeInfoDbData_1.2.10
[71] cli_3.6.1              
[72] fansi_1.0.5            
[73] gtable_0.3.4           
[74] digest_0.6.33          
[75] ggrepel_0.9.4          
[76] memoise_2.0.1          
[77] htmltools_0.5.6.1      
[78] lifecycle_1.0.3        
[79] httr_1.4.7             
[80] mime_0.12              
[81] MASS_7.3-60            
[82] bit64_4.0.5
```


# Read in data
## Read in counts

- need to make into a nice table with each sample as a column header that makes sense
- col_types = "n" specifies that the data is numeric
```{r echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
setwd("data_in/Counts")
f <- c("TranscriptID", list.files(pattern = "*.tsv") %>% str_remove("_.*"))

f1 <- c("blank", list.files(pattern = "*.tsv"))

dat <- read_tsv(f1[2], col_names = f[1:2], comment = "_")

for (i in 3:length(f1)){
  dat <- full_join(dat, read_tsv(f1[i], col_names = f[c(1,i)], comment = "_"))
}

tid <- dat$TranscriptID
dat <- dat %>% dplyr::select(!TranscriptID)
dat <- as.matrix(dat)

rownames(dat) <- tid

as.data.frame(head(dat))
```

## Read in metadata
```{r}
meta <- read_csv("data_in/metadata.csv") %>% mutate(Group = factor(paste(Treatment,Time, sep = "")))
meta
```

# Convert to DGEList object
This is an edgeR-specific data type which holds all the information needed for any analysis
```{r}
y <- DGEList(counts=dat, group=meta$Group)

# normalize
y <- calcNormFactors(y, method="TMM")

y$samples
```

## Make a matrix indicating the relationship between different samples
```{r}
design <- model.matrix(~0+meta$Group, data = y$samples)
colnames(design) <- levels(y$samples$group)
design
```

## Filter lowly expressed genes

Only counts with at least 30 reads will be kept
```{r}
keep <- filterByExpr(y, min.count = 30, design = design, group = meta$Group)
y <- y[keep,,keep.lib.sizes=FALSE]
head(y$counts)
```

## Estimate dispersion

- Can be done with linear model or generalized linear model (GLM)
  - **GLM is more flexible, so I will be using this**
- Common dispersion = mean dispersion across all genes
- Trended dispersion = mean dispersion across all genes with similar abundance (the fitted value of the mean-dispersion trend)
- Tagwise dispersion = "raw" tagwise estimates (not information across genes)
- Too much variability will mean DE analysis will have a lot of noise
- These estimates help ensure the correct model is used for DE analysis later
- "Biological CV (BCV) is the coefficient of variation (CV) with which the (unknown) true abundance of the gene varies between replicate RNA samples. **It represents the CV that would remain between biological replicates if sequencing depth could be increased indefinitely.** The **technical CV decreases as the size of the counts increases. BCV on the other hand does not.** BCV is therefore likely to be the dominant source of uncertainty for high-count genes, so reliable estimation of BCV is crucial for realistic assessment of differential expression in RNA-Seq experiments. If the abundance of each gene varies between replicate RNA samples in such a way that the genewise standard deviations are proportional to the genewise means, a commonly occurring property of measurements on physical quantities, then it is reasonable to suppose that BCV is approximately constant across genes. We allow however for the possibility that BCV might vary between genes and might also show a systematic trend with respect to gene expression or expected count.The magnitude of BCV is more important than the exact probabilistic law followed by the true gene abundances. For mathematical convenience, we assume that the true gene abundances follow a gamma distributional law between replicate RNA samples. This implies that the read counts follow a negative binomial probability law." (https://academic.oup.com/nar/article/40/10/4288/2411520)
- "Common dispersion (i.e. red line on the BCV plot) between **0.2 and 0.4 is usually considered reasonable** and hence could detect more DE genes. If the common dispersion is above the 0.4 threshold, this will influence the number of DE genes found in the study." (https://bioinformatics.cvr.ac.uk/some-key-factors-for-number-of-significant-de-genes/)
  - Us having common dispersion close to 0 means that the overall variation between times, replicates, and treatments is low, which will mean fewer differentially expressed genes being detected\

```{r}
#regular dispersion
y <- estimateDisp(y,design)
# estimate common dispersion:
y <- estimateCommonDisp(y)
# to estimate tagwise dispersions:
y <- estimateTagwiseDisp(y)

plotBCV(y, main = "Regular dispersion")
```

Note: only the trended dispersion is used under the quasi-likelihood (QL) pipeline
```{r}
# GLM dispersion
# estimate common dispersion
y <- estimateGLMCommonDisp(y, design)
# estimate trended dispersions:
y <- estimateGLMTrendedDisp(y, design)
# estimate tagwise dispersions:
y <- estimateGLMTagwiseDisp(y, design)

y

plotBCV(y, main = "GLM dispersion")
```

## Mean-difference plot (MD-plot)

- What is plotMD? "A mean-difference plot (MD-plot) is a plot of log fold changes (differences) versus average log values (means). The history of mean-difference plots and MA-plots is reviewed in Ritchie et al (2015)." (https://rdrr.io/bioc/edgeR/man/plotMD.html)
- "The performance of the TMM normalization procedure can be examined using mean-difference (MD) plots. This **visualizes the library size-adjusted log-fold change between two libraries (the difference) against the average log-expression across those libraries (the mean)**. The following MD plot is generated by comparing sample 1 against an artificial library constructed from the average of all other samples." ... "**Ideally, the bulk of genes should be centred at a log-fold change of zero. This indicates that any composition bias between libraries has been successfully removed.** This quality check should be repeated by constructing a MD plot for each sample."

```{r eval=F}
for (i in 1:30){
  plotMD(cpm(y, log=TRUE), column=i)
  abline(h=0, col="red", lty=2, lwd=2)
}
```

##  Multi-dimensional scaling (MDS) plot

What is MDS plot? Visualization of the differences between the expression profiles of different samples in two dimensions. It does this by calculating the relative distance between samples as a distance matrix.

Note: this is Supplementary Figure S4
```{r fig.width = 6, fig.height = 6.5}
points <- c(0,1,2,3,4,15,16,17,18,19)
colors <- rep(c("#b37d36", "#36b37e", "#3668b3", "#6e36b3", "#b33670"), 3)
plotMDS(y, col=colors[meta$Group], pch=points[meta$Group])
legend("topleft", legend=levels(meta$Group), pch=points, col=colors, ncol=2)
```


# Do differential expression analysis

These are GLM-based tests: glmQLFit() fits the negative binomial

GLM for each tag
- QL = quasi-likelihood
- Raw = estimated values of the GLM coefficients for each gene
- Trend = fitted mean-QL dispersion trend
- Squeezed = squeezed QL estimates
  - This method is adapted from this paper: http://www.statsci.org/smyth/pubs/QuasiSeqPreprint.pdf
  - "Estimation of the gene-specific QL dispersion is difficult as most RNA-seq data sets have limited numbers of replicates. This means that there is often little information to stably estimate the dispersion for each gene. To overcome this, an empirical Bayes (EB) approach is used whereby information is shared between genes [16, 7, 17]. Briefly, a mean-dependent trend is fitted to the raw QL dispersion estimates. The raw estimates are then squeezed towards this trend to obtain moderated EB estimates, which can be used in place of the raw values for downstream hypothesis testing. This EB strategy reduces the uncertainty of the estimates and improves testing power." http://www.statsci.org/smyth/pubs/QLedgeRPreprint.pdf

```{r}
# QL dispersion estimation and hypothesis testing
fit <- glmQLFit(y,design, robust = T)

plotQLDisp(fit, main = "QL dispersion estimation")
```

# Prepare contrasts

```{r}
my.contrasts <- makeContrasts(
  JAvsEt1 = (JA1h-JA0h)-(Et1h-Et0h),
  JAvsEt2 = (JA2h-JA0h)-(Et2h-Et0h),
  JAvsEt4 = (JA4h-JA0h)-(Et4h-Et0h),
  JAvsEt6 = (JA6h-JA0h)-(Et6h-Et0h), 
  levels = design)
```

# Do differential expression analysis taking taking into account minimum FC 1.5

- Cutoffs are FDR <0.01 and fold change > 1.5 
- There is no mathematical reason that a gene below any fold change cutoff wouldn't be statistically significantly differentially expressed, but I opted for a small cutoff for biological reasons
- Page 5 of egdeR manual: "Note that the fold-change threshold in glmTreat() is not the minimum value of the fold-change expected to see from the testing results. Genes will need to exceed this threshold by some way before being declared statistically significant. It is better to interpret the threshold as “the fold-change below which we are definitely not interested in the gene" rather than “the fold-change above which we are interested in the gene". "

```{r}
lfc = log2(1.5)
# do DE for 1h
qlf <- glmTreat(fit, contrast = my.contrasts[,"JAvsEt1"], lfc=lfc)

# calculate FDR for all
pairwise_de <- as.data.frame(topTags(qlf, n = nrow(dat), sort.by = "none")) 
pairwise_de <- pairwise_de %>% 
  dplyr::select(c(logFC, logCPM, FDR)) %>% 
  dplyr::rename("logFC_JAvsEt1" = logFC, "logCPM_JAvsEt1" = logCPM, "FDR_JAvsEt1" = FDR) %>% 
  rownames_to_column(var = "TranscriptId")

# do DE for 2h
qlf <- glmTreat(fit, contrast = my.contrasts[,"JAvsEt2"], lfc=lfc)

# calculate FDR
temp <- as.data.frame(topTags(qlf, n = nrow(dat), sort.by = "none")) %>% 
  dplyr::select(c(logFC, logCPM, FDR)) %>% 
  dplyr::rename("logFC_JAvsEt2" = logFC, "logCPM_JAvsEt2" = logCPM, "FDR_JAvsEt2" = FDR) %>% 
  rownames_to_column(var = "TranscriptId")

# merge
pairwise_de <- merge(pairwise_de, temp, by="TranscriptId")

# do DE for 4h
qlf <- glmTreat(fit, contrast = my.contrasts[,"JAvsEt4"], lfc=lfc)

# calculate FDR
temp <- as.data.frame(topTags(qlf, n = nrow(dat), sort.by = "none")) %>% 
  dplyr::select(c(logFC, logCPM, FDR)) %>% 
  dplyr::rename("logFC_JAvsEt4" = logFC, "logCPM_JAvsEt4" = logCPM, "FDR_JAvsEt4" = FDR) %>% 
  rownames_to_column(var = "TranscriptId")

# merge
pairwise_de <- merge(pairwise_de, temp, by="TranscriptId")

# do DE for 6h
qlf <- glmTreat(fit, contrast = my.contrasts[,"JAvsEt6"], lfc=lfc)

# calculate FDR
temp <- as.data.frame(topTags(qlf, n = nrow(dat), sort.by = "none")) %>% 
  dplyr::select(c(logFC, logCPM, FDR)) %>% 
  dplyr::rename("logFC_JAvsEt6" = logFC, "logCPM_JAvsEt6" = logCPM, "FDR_JAvsEt6" = FDR) %>% 
  rownames_to_column(var = "TranscriptId")

# merge
pairwise_de <- merge(pairwise_de, temp, by="TranscriptId")

pairwise_de
```

Write to file

```{r}
write.csv(pairwise_de, "data_out/Supporting_Data_S2.csv", row.names = F)
```

## Select differentially expressed genes (FDR < 0.01 in at least one pairwise comparison)

```{r}
fdr = 0.01
de_genes <- pairwise_de %>% 
  filter(FDR_JAvsEt1 < fdr | 
           FDR_JAvsEt2 < fdr | 
           FDR_JAvsEt4 < fdr | 
           FDR_JAvsEt6 < fdr) %>% 
  dplyr::select(c(TranscriptId, logFC_JAvsEt1, logFC_JAvsEt2, logFC_JAvsEt4, logFC_JAvsEt6))

de_genes
```

# Cluster analysis

### Data before scaling

```{r}
scaledata_fc <- de_genes %>% column_to_rownames(var = "TranscriptId")
paste("Unscaled data range: ", round(min(scaledata_fc),2), "-", round(max(scaledata_fc),2))
paste("Number of transcripts before scaling:", nrow(scaledata_fc))
```


### Scale and center log2FC

```{r}
scaledata_fc <- t(scale(t(as.matrix(scaledata_fc)), center = T))
# filter out missing data (can result from scaling)
scaledata_fc <- scaledata_fc[complete.cases(scaledata_fc),]
head(as.data.frame(scaledata_fc), n=10)
paste("Scaled data range: ", round(min(scaledata_fc),2), "-", round(max(scaledata_fc),2))
paste("Number of transcripts after scaling:", nrow(scaledata_fc))
```

## dtwclust

### Convert to list format

```{r}
set.seed(124)
scaledata_t <- scaledata_fc
colnames(scaledata_t) <- c(1,2,4,6)
scaledata_t <- tslist(scaledata_t, simplify=T)
head(scaledata_t)
```

### Calculate clusters

Three algorithms with default settings and different numbers of clusters:

- Algorithms: partitional, hierarchical, tadpole
- Cluster sizes: 4, 6, 8, 10

```{r}
cluster_part_k4 <- tsclust(series = scaledata_t, type = "partitional", k = 4, preproc = NULL, distance = "dtw_basic", centroid = "pam", args = tsclust_args(), seed = 1245, trace = T, error.check = T)
cluster_part_k6 <- tsclust(series = scaledata_t, type = "partitional", k = 6, preproc = NULL, distance = "dtw_basic", centroid = "pam", args = tsclust_args(), seed = 1245, trace = T, error.check = T)
cluster_part_k8 <- tsclust(series = scaledata_t, type = "partitional", k = 8, preproc = NULL, distance = "dtw_basic", centroid = "pam", args = tsclust_args(), seed = 1245, trace = T, error.check = T)
cluster_part_k10 <- tsclust(series = scaledata_t, type = "partitional", k = 10, preproc = NULL, distance = "dtw_basic", centroid = "pam", args = tsclust_args(), seed = 1245, trace = T, error.check = T)

cluster_hi_k4 <- tsclust(series = scaledata_t, type = "hierarchical", k = 4, preproc = NULL, distance = "dtw_basic", centroid = "pam", args = tsclust_args(), seed = 1245, trace = T, error.check = T)
cluster_hi_k6 <- tsclust(series = scaledata_t, type = "hierarchical", k = 6, preproc = NULL, distance = "dtw_basic", centroid = "pam", args = tsclust_args(), seed = 1245, trace = T, error.check = T)
cluster_hi_k8 <- tsclust(series = scaledata_t, type = "hierarchical", k = 8, preproc = NULL, distance = "dtw_basic", centroid = "pam", args = tsclust_args(), seed = 1245, trace = T, error.check = T)
cluster_hi_k10 <- tsclust(series = scaledata_t, type = "hierarchical", k = 10, preproc = NULL, distance = "dtw_basic", centroid = "pam", args = tsclust_args(), seed = 1245, trace = T, error.check = T)

cluster_tad_k4 <- tsclust(series = scaledata_t, type = "tadpole", k = 4, preproc = NULL, args = tsclust_args(), seed = 1245, trace = T, error.check = T, control = tadpole_control(dc = 1, window.size = 1))
cluster_tad_k6 <- tsclust(series = scaledata_t, type = "tadpole", k = 6, preproc = NULL, args = tsclust_args(), seed = 1245, trace = T, error.check = T, control = tadpole_control(dc = 1, window.size = 1))
cluster_tad_k8 <- tsclust(series = scaledata_t, type = "tadpole", k = 8, preproc = NULL, args = tsclust_args(), seed = 1245, trace = T, error.check = T, control = tadpole_control(dc = 1, window.size = 1))
cluster_tad_k10 <- tsclust(series = scaledata_t, type = "tadpole", k = 10, preproc = NULL, args = tsclust_args(), seed = 1245, trace = T, error.check = T, control = tadpole_control(dc = 1, window.size = 1))
```

Crisp partitions

– "Sil" (!): Silhouette index (Rousseeuw (1987); to be maximized).
– "D" (!): Dunn index (Arbelaitz et al. (2013); to be maximized).
– "COP" (!): COP index (Arbelaitz et al. (2013); to be minimized).
– "DB" (?): Davies-Bouldin index (Arbelaitz et al. (2013); to be minimized).
– "DBstar" (?): Modified Davies-Bouldin index (DB*) (Kim and Ramakrishna (2005); to be minimized).
– "CH" (~): Calinski-Harabasz index (Arbelaitz et al. (2013); to be maximized).
– "SF" (~): Score Function (Saitta et al. (2007); to be maximized; see notes).

Fuzzy partitions (using the nomenclature from Wang and Zhang (2007))

– "MPC": to be maximized.
– "K" (~): to be minimized.
– "T": to be minimized.
– "SC" (~): to be maximized.
– "PBMF" (~): to be maximized (see notes)
```{r}
best_cluster <- as.data.frame(cvi(cluster_part_k4)) %>%
  cbind(as.data.frame(cvi(cluster_part_k6))) %>%
  cbind(as.data.frame(cvi(cluster_part_k8))) %>% 
  cbind(as.data.frame(cvi(cluster_part_k10))) %>% 
  cbind(as.data.frame(cvi(cluster_hi_k4))) %>%
  cbind(as.data.frame(cvi(cluster_hi_k6))) %>%
  cbind(as.data.frame(cvi(cluster_hi_k8))) %>%
  cbind(as.data.frame(cvi(cluster_tad_k4))) %>%
  cbind(as.data.frame(cvi(cluster_tad_k6))) %>% 
  cbind(as.data.frame(cvi(cluster_tad_k8))) %>%
  cbind(as.data.frame(cvi(cluster_tad_k10))) %>% 
  mutate(desired_outcome = c("max", "max", "max", "min", "min", "max", "min"))

#display the algorithm with the max score, for those metrics that want that
best_cluster$result[best_cluster$desired_outcome == "max"] <- colnames(best_cluster[best_cluster$desired_outcome == "max",])[apply(best_cluster[best_cluster$desired_outcome == "max",],1,which.max)]

#display the algorithm with the min score, for those metrics that want that
best_cluster$result[best_cluster$desired_outcome == "min"] <- colnames(best_cluster[best_cluster$desired_outcome == "min",])[apply(best_cluster[best_cluster$desired_outcome == "min",],1,which.min)]

best_cluster
as.data.frame(table(best_cluster$result))
```

Hierarchical is best, now see which distance method is best: dtw_basic or euclidean
```{r}
cluster_hi_e_k4 <- tsclust(series = scaledata_t, type = "hierarchical", k = 4, preproc = NULL, distance = "euclidean", centroid = "pam", args = tsclust_args(), seed = 1245, trace = T, error.check = T)
```

```{r}
best_cluster <- as.data.frame(cvi(cluster_hi_k4)) %>%
  cbind(as.data.frame(cvi(cluster_hi_e_k4))) %>%
  mutate(desired_outcome = c("max", "max", "max", "min", "min", "max", "min"))

#display the algorithm with the max score, for those metrics that want that
best_cluster$result[best_cluster$desired_outcome == "max"] <- colnames(best_cluster[best_cluster$desired_outcome == "max",])[apply(best_cluster[best_cluster$desired_outcome == "max",],1,which.max)]

#display the algorithm with the min score, for those metrics that want that
best_cluster$result[best_cluster$desired_outcome == "min"] <- colnames(best_cluster[best_cluster$desired_outcome == "min",])[apply(best_cluster[best_cluster$desired_outcome == "min",],1,which.min)]

best_cluster
as.data.frame(table(best_cluster$result))
```

Euclidean is best, now see which control is best: “ward.D2”, “single”, “average”, “mcquitty”, “centroid”
```{r}
cluster_hi_k4s <- tsclust(series = scaledata_t, type = "hierarchical", k = 4, preproc = NULL, distance = "euclidean", args = tsclust_args(), seed = 1245, trace = T, error.check = T, control = hierarchical_control(method = "single"))
cluster_hi_k4w <- tsclust(series = scaledata_t, type = "hierarchical", k = 4, preproc = NULL, distance = "euclidean", args = tsclust_args(), seed = 1245, trace = T, error.check = T, control = hierarchical_control(method = "ward.D2"))
cluster_hi_k4c <- tsclust(series = scaledata_t, type = "hierarchical", k = 4, preproc = NULL, distance = "euclidean", args = tsclust_args(), seed = 1245, trace = T, error.check = T, control = hierarchical_control(method = "centroid"))
cluster_hi_k4a <- tsclust(series = scaledata_t, type = "hierarchical", k = 4, preproc = NULL, distance = "euclidean", args = tsclust_args(), seed = 1245, trace = T, error.check = T, control = hierarchical_control(method = "average"))
cluster_hi_k4m <- tsclust(series = scaledata_t, type = "hierarchical", k = 4, preproc = NULL, distance = "euclidean", args = tsclust_args(), seed = 1245, trace = T, error.check = T, control = hierarchical_control(method = "mcquitty"))
```

If you want to explore other options in real time (I tried some larger clusters just in case)
```
interactive_clustering(scaledata_t)
```

```{r}
best_cluster <- as.data.frame(cvi(cluster_hi_k4s)) %>%
  cbind(as.data.frame(cvi(cluster_hi_k4w))) %>%
  cbind(as.data.frame(cvi(cluster_hi_k4c))) %>% 
  cbind(as.data.frame(cvi(cluster_hi_k4a))) %>% 
  cbind(as.data.frame(cvi(cluster_hi_k4m))) %>%
  mutate(desired_outcome = c("max", "max", "max", "min", "min", "max", "min"))

#display the algorithm with the max score, for those metrics that want that
best_cluster$result[best_cluster$desired_outcome == "max"] <- colnames(best_cluster[best_cluster$desired_outcome == "max",])[apply(best_cluster[best_cluster$desired_outcome == "max",],1,which.max)]

#display the algorithm with the min score, for those metrics that want that
best_cluster$result[best_cluster$desired_outcome == "min"] <- colnames(best_cluster[best_cluster$desired_outcome == "min",])[apply(best_cluster[best_cluster$desired_outcome == "min",],1,which.min)]

best_cluster
```


## Chose cluster

```{r}
cluster_final <- cluster_hi_k4c
```

## Make table

```{r}
time_cluster <- data.frame("TranscriptId" = names(cluster_final@datalist), Cluster = cluster_final@cluster)
time_cluster
```

### How many genes per cluster?

```{r fig.width = 5, fig.height = 2}
colors <- c("#CC4C02", "#35254AFF", "#3493A8FF", "#FEC44F")

names(colors) <- as.character(1:4)
plot(seq_len(length(colors)), 
     rep_len(1, length(colors)),
     col = colors, 
     pch = 16, 
     cex = 6, 
     xaxt = 'n', 
     yaxt = 'n', 
     xlab = '', 
     ylab = '')
```

```{r}
an <- as.data.frame(table(time_cluster$Cluster)) %>% rename(Var1 = "Cluster", Freq = "Num_of_genes") %>% 
  ggplot(.) +
  aes(x = Cluster, y = Num_of_genes, fill = Cluster) +
  geom_col() +
  scale_fill_manual(values = colors) +
  guides(fill = "none") +
  theme_bw() +
  labs(y = "Number of Genes") +
  scale_y_continuous(breaks = seq(0 , 3000, 500), expand = c(0.005,0.005))
an
```

# Interpro annotations

https://interproscan-docs.readthedocs.io/en/latest/OutputFormats.html

Available analyses:

- TIGRFAM : TIGRFAMs are protein families based on hidden Markov models (HMMs).
- SFLD : SFLD is a database of protein families based on hidden Markov models (HMMs).
- SUPERFAMILY : SUPERFAMILY is a database of structural and functional annotations for all proteins and genomes.
- PANTHER : The PANTHER (Protein ANalysis THrough Evolutionary Relationships) Classification System is a unique resource that classifies genes by their functions, using published scientific experimental evidence and evolutionary relationships to predict function even in the absence of direct experimental evidence.
- Gene3D : Structural assignment for whole genes and genomes using the CATH domain structure database.
- Hamap : High-quality Automated and Manual Annotation of Microbial Proteomes.
- ProSiteProfiles : PROSITE consists of documentation entries describing protein domains, families and functional sites as well as associated patterns and profiles to identify them.
- Coils : Prediction of coiled coil regions in proteins.
- SMART : SMART allows the identification and analysis of domain architectures based on hidden Markov models (HMMs).
- CDD : CDD predicts protein domains and families based on a collection of well-annotated multiple sequence alignment models.
- PRINTS : A compendium of protein fingerprints - a fingerprint is a group of conserved motifs used to characterise a protein family.
- PIRSR : PIRSR is a database of protein families based on hidden Markov models (HMMs) and Site Rules.
- ProSitePatterns : PROSITE consists of documentation entries describing protein domains, families and functional sites as well as associated patterns and profiles to identify them.
- AntiFam : AntiFam is a resource of profile-HMMs designed to identify spurious protein predictions.
- Pfam : A large collection of protein families, each represented by multiple sequence alignments and hidden Markov models (HMMs).
- MobiDBLite : Prediction of intrinsically disordered regions in proteins.
- PIRSF : The PIRSF concept is used as a guiding principle to provide comprehensive and non-overlapping clustering of UniProtKB sequences into a hierarchical order to reflect their evolutionary relationships.

In the interpro table:

- Column 1: the protein identifier
- Column 5: the identifier of the signature that was found in the protein sequence
- Column 4: the databank where this signature comes from (InterProScan regroups several motifs databanks)
- Column 6: the human readable description of the motif
- Columns 7 and 8: the position where the motif was found
- Column 9: a score for the match (if available)
- Column 12 and 13: identifier of the signature integrated in InterPro (if available). Have a look an example webpage for IPR036859 on InterPro.
- https://training.galaxyproject.org/training-material/topics/genome-annotation/tutorials/functional/tutorial.html

```{r}
interpro <- read_tsv("data_in/annotations_combined_pep.fa.tsv", col_names = c("Protein_accession", "Sequence_MD5_digest", "Sequence_length", "Analysis", "Signature_accession", "Signature_description", "Start_location", "Stop_location", "Evalue", "Status_of_match", "Date", "InterPro_annotations", "InterPro_annotation_description", "GO_annotations", "Pathways_annotations")) %>% 
  mutate(Protein_accession = str_remove(Protein_accession, "_[0-9]")) %>% 
  dplyr::select(!c(Sequence_MD5_digest,Pathways_annotations))

head(interpro)
```

How many genes?

```{r}
length(unique(interpro$Protein_accession))
```

Distribution of genes per contig

```{r}
gff3 <- read_tsv("data_in/annotations_combined.gff3", col_names = NULL, comment = "#")
filter(gff3, X3 == "transcript")

top_contigs <- read_delim("data_in/seq_lengths.tsv", delim = " ", col_names = F) %>% 
  arrange(desc(X2))
top_contigs
```

Output:
```
ptg000001l	51110158		
ptg000033l	32623551		
ptg000031l	29529208		
ptg000053l	28467466		
ptg000005l	25785398		
ptg000019l	25264053		
ptg000009l	24533512		
ptg000014l	23505251		
ptg000030l	23301990		
ptg000036l	22950366
...
```

```{r}
paste("Total genes:", nrow(filter(gff3, X3 == "transcript")))

paste("Number of genes in top 100 largest contigs:", sum(str_detect(gff3$X1[gff3$X3 == "transcript"], paste0(top_contigs$X1[1:100], collapse = "|")), na.rm = T))

paste("Number of genes in top 51 largest contigs (this is the L75 number):", sum(str_detect(gff3$X1[gff3$X3 == "transcript"], paste0(top_contigs$X1[1:51], collapse = "|")), na.rm = T))

paste("Number of genes in top 46 largest contigs (this is the L50 number):", sum(str_detect(gff3$X1[gff3$X3 == "transcript"], paste0(top_contigs$X1[1:26], collapse = "|")), na.rm = T))
```

Output
```
[1] "Total genes: 34107"
[1] "Number of genes in top 100 largest contigs: 33410"
[1] "Number of genes in top 51 largest contigs (this is the L75 number): 27667"
[1] "Number of genes in top 46 largest contigs (this is the L50 number): 18332"
```

# Refine GO annotations

## Add blast results against swissprot

```{r}
swi <- read_csv("data_in/blast_swiss_annotations_combined.csv", col_names = NULL) %>% 
  # select only top hit
  # group by transcript id
  group_by(X1) %>% 
  # select hit with lowest evalue
  dplyr::slice(which.min(X11)) %>% 
  # select the transcrtipt id and associated swissprot accession
  dplyr::select(c(X1, X2)) %>% 
  dplyr::rename("Protein_accession" = X1, "Entry" = X2) %>% 
  mutate(Protein_accession = str_remove(Protein_accession, "_[0-9]"))
head(swi)
```

```{r}
write(paste(unique(swi$Entry), collapse = " "), "data_out/swissprot_hits.txt")
```

1. Go to https://www.uniprot.org/id-mapping and load the text file
2. Select from: UniProtKB_AC-ID     to: UniProtKB-Swiss-Prot
3. Select the results when loaded
4. Click 'customize columns' deselect all defaults, and select 'gene ontology IDs', "gene names", and "descriptions"
5. Download results as tsv, uncompressed
6. Rename to "get_swissprot_go.tsv"

```{r}
temp <- read_tsv("data_in/get_swissprot_go.tsv") %>% dplyr::select(c(From, Gene_Ontology_IDs)) %>% dplyr::rename("Entry" = From, "GO_sw" = "Gene_Ontology_IDs")

head(temp)
#check for improper concatenation
temp %>% filter(str_detect(GO_sw, "[0-9]GO"))
```

```{r}
swi <- full_join(swi, temp, by = "Entry") %>% mutate(GO_sw = str_replace_all(GO_sw, "; ", ",")) %>% dplyr::select(!Entry) %>% mutate(GO_sw = ifelse(is.na(GO_sw), "", GO_sw))
swi
#check for improper concatenation
swi %>% filter(str_detect(GO_sw, "[0-9]GO"))
```

## Get GO terms per transcript

```{r}
GO_clean <- interpro[,c(1,13)] %>% 
  # group by transcriptID (as each has multiple lines)
  group_by(Protein_accession) %>% 
  # add a new column concatenating all that are in the same group, and remove the old GO column
  mutate(GO = paste0(GO_annotations, collapse = ","), .keep = "unused") %>% 
  # the previous step means that each duplicate will contain all the same data in the GO column, so now can delete duplicates
  distinct(Protein_accession, .keep_all = T) %>% 
  # clean up GO column to delete "NA", "-", and "|"
  mutate(GO = str_remove_all(GO, "NA[,]*")) %>% 
  mutate(GO = str_remove_all(GO, "-[,]*")) %>% 
  mutate(GO = str_remove_all(GO, ",$")) %>% 
  mutate(GO = str_replace_all(GO, "\\|", ","))
GO_clean
#check for improper concatenation
GO_clean %>% filter(str_detect(GO, "[0-9]GO"))
```

### Combine with swissprot

```{r}
GO_clean <- full_join(GO_clean, swi, by = "Protein_accession") %>% 
  # convert added NAs to blanks
  replace(is.na(.), "")
GO_clean <- GO_clean %>% 
  unite("GO", c(GO, GO_sw), sep = ",", remove = T) %>% 
  # remove extra commas
  # doubles
  mutate(GO = str_replace(GO, ",,", ",")) %>% 
  # at start
  mutate(GO = str_remove(GO, "^,")) %>% 
  # at end
  mutate(GO = str_remove(GO, ",$")) %>% 
  # remove go_sw now that part of GO
  #dplyr::select(c(Protein_accession, GO)) %>% 
  # split cleaned GO terms for removal of duplicates (next step)
  mutate(GO_str = strsplit(GO, ","))
GO_clean
#check for improper concatenation
GO_clean %>% filter(str_detect(GO, "[0-9]GO"))
```

### Get rid of duplicate GO annotations

```{r}
for (i in 1:nrow(GO_clean)){
  GO_clean$GO[i] <- paste0(unique(unlist(GO_clean$GO_str[i])), collapse = ",")
}

GO_clean <- GO_clean %>% 
  # convert blanks to NAs
  mutate_all(na_if,"") %>% 
  # remove GO_str
  dplyr::select(!GO_str)
GO_clean
```

### Check go annotations

```{r}
GO_names <- GO_clean$GO[!is.na(GO_clean$GO)]
GO_names <- paste0(GO_names, collapse = ",")
GO_names <- strsplit(GO_names, ",")
GO_names <- as.data.frame(GO_names, col.names = "GO")
GO_names <- as.data.frame(Term(GOTERM)) %>% rename("Term(GOTERM)" = "GO_term") %>% rownames_to_column(var = "GO") %>% left_join(GO_names, ., by = "GO") %>% distinct(GO, .keep_all = T) 
GO_names
```

Look through for keywords that don't belong

```{r}
bad_GO <- GO_names %>% filter(!is.na(GO_term)) %>% filter(str_detect(GO_term, "aging|larval|nervous system|neur[oa]|behavior|mating|animal|glial|imaginal disc|wing|Schwann|cardiac|atrial|kidney|intestine|stomach|eye|retina|spleen|axon |myoblast|bone|osteo|male|MHC"))
bad_GO
```

Where are these incorrect GO terms coming from (as in, which prediction tool)?

```{r}
swi %>% filter(str_detect(GO_sw, paste0(bad_GO$GO, collapse = "|")))
interpro[,c(1,13)] %>% 
  # group by transcriptID (as each has multiple lines)
  group_by(Protein_accession) %>% 
  # add a new column concatenating all that are in the same group, and remove the old GO column
  mutate(GO = paste0(GO_annotations, collapse = ","), .keep = "unused") %>% 
  # the previous step means that each duplicate will contain all the same data in the GO column, so now can delete duplicates
  distinct(Protein_accession, .keep_all = T) %>% 
  # clean up GO column to delete "NA", "-", and "|"
  mutate(GO = str_remove_all(GO, "NA[,]*")) %>% 
  mutate(GO = str_remove_all(GO, "-[,]*")) %>% 
  mutate(GO = str_remove_all(GO, ",$")) %>% 
  mutate(GO = str_replace_all(GO, "\\|", ",")) %>% 
  filter(str_detect(GO, paste0(bad_GO$GO, collapse = "|")))
```

#### Remove problematic GO terms

```{r}
GO_clean <- GO_clean %>% 
  # remove incorrect GO terms
  mutate(GO = str_remove_all(GO, paste0(bad_GO$GO, collapse = "|"))) %>% 
  # remove multiple commas
  mutate(GO = str_replace_all(GO, ",[,]+", ","))
GO_clean
```

```{r}
write_tsv(GO_clean, "data_out/Supporting_Data_S1.tsv", na = "", col_names = F)
```

# GO analysis

```{r}
geneID2GO <- readMappings(file = "data_out/Supporting_Data_S1.tsv")
str(head(geneID2GO))
geneNames <- names(geneID2GO)
```

"Tests based on gene counts. This is the most popular family of tests, given that it only requires the presence of a list of interesting genes and nothing more. Tests like Fisher's exact test, Hypegeometric test and binomial test belong to this family."

Why is there no FDR? Would result in too conservative an estimate (hardly any GO terms), doesn't take into account all the variables that went into determining the p-value (eg. go topology), and because the go terms aren't independent this kind of testing doesn't meet the assumptions of multiple testing correction

https://bioconductor.org/packages/devel/bioc/vignettes/topGO/inst/doc/topGO.pdf

```{r}
whichTests()
whichAlgorithms()
```

## Cluster from dtwclust

```{r}
sig_go_terms_t <- data.frame(GO.ID = character(), Term = character(), Annotated = integer(), Significant = integer(), Expected = double(), "Rank in classic" = integer(), classic = character(), elim = character(), weight = character(), Cluster = character())

for (i in 1:length(unique(time_cluster$Cluster))) {
  myInterestingGenes <- time_cluster$TranscriptId[time_cluster$Cluster == i]
  geneList <- factor(as.integer(geneNames %in% myInterestingGenes))
  names(geneList) <- geneNames
  
  GOdata <- new("topGOdata", ontology = "BP", allGenes = geneList, annot = annFUN.gene2GO, gene2GO = geneID2GO, nodeSize = 8)
  
  resultFis <- runTest(GOdata, algorithm = "classic", statistic = "fisher")
  resultFisher <- runTest(GOdata, algorithm = "elim", statistic = "fisher")
  resultWeight <- runTest(GOdata, algorithm = "weight", statistic = "fisher")
  
  allRes <- GenTable(GOdata, classic = resultFis, elim = resultFisher, weight = resultWeight, orderBy = "weight", ranksOf = "classic", topNodes = 25)
  allRes$Cluster <- paste0("Cluster", i)
  
  sig_go_terms_t <- rbind(sig_go_terms_t, allRes)
}

sig_go_terms <- sig_go_terms_t %>% 
  mutate(classic = as.numeric(classic)) %>% 
  mutate(elim = as.numeric(elim)) %>% 
  mutate(weight = as.numeric(weight)) %>% 
  filter(elim < 0.01) %>% 
  filter(weight < 0.01) %>% 
  filter(classic < 0.01)

sig_go_terms
```

```{r}
write.csv(sig_go_terms, "data_out/Supporting_Data_S5.csv", row.names = F)
```

## Make plots

```{r fig.width = 5, fig.height = 2}
colors <- c("#CC4C02", "#35254AFF", "#3493A8FF", "#FEC44F")

names(colors) <- as.character(1:4)
plot(seq_len(length(colors)), 
     rep_len(1, length(colors)),
     col = colors, 
     pch = 16, 
     cex = 6, 
     xaxt = 'n', 
     yaxt = 'n', 
     xlab = '', 
     ylab = '')
```

```{r}
x <- sig_go_terms %>% 
  group_by(Cluster) %>% 
  mutate(Percentage_of_total = (Significant/Annotated)*100) %>% 
  mutate(Term = str_replace(Term, "biosyn.*", "biosynthesis")) %>% 
  mutate(Term = str_replace(Term, "metabol.*", "metabolism")) %>% 
  mutate(Term = str_replace(Term, "involved.*", "")) %>% 
  mutate(Term = str_replace(Term, "intracellular.*", "transport")) %>% 
  mutate(Term = str_replace(Term, "via b.*", "transport"))
```

```{r fig.width = 9, fig.height = 8}
plota <- x %>%
 mutate(Cluster = str_extract(Cluster, "[0-9]+")) %>% 
 mutate(Term = fct_reorder(Term, Cluster)) %>% 
 ggplot() +
  aes(x = Percentage_of_total, y = Term, fill = Cluster) +
  geom_col() +
  scale_fill_manual(values = colors) +
  labs(x = "GO term representation\nwithin cluster (%)", size = "Count in cluster", y = "GO term") +
  theme_bw() +
  theme(legend.position="bottom") +
  guides(fill = "none") +
  scale_x_continuous(breaks = seq(0 , 80, 5), expand = c(0.005,0.005), limits = c(0,75)) +
  theme(panel.grid.major.y = element_blank(), text = element_text(size = 12)) +
  geom_vline(xintercept = c(10, 20, 30, 40, 50, 60), color = "grey", alpha = 0.4)

plotb <- as.data.frame(table(time_cluster$Cluster)) %>% rename(Var1 = "Cluster", Freq = "Num_of_genes") %>% 
  ggplot(.) +
  aes(x = Cluster, y = Num_of_genes, fill = Cluster) +
  geom_col() +
  scale_fill_manual(values = colors) +
  guides(fill = "none") +
  theme_bw() +
  labs(y = "Genes") +
  scale_y_continuous(breaks = seq(0 , 3000, 500), expand = c(0.005,0.005))

plotc <- de_genes %>% right_join(time_cluster,., by = "TranscriptId") %>% 
  pivot_longer(3:ncol(.)) %>% 
  mutate(time = str_extract(name, "[0-9]")) %>% 
  mutate(Cluster = as.factor(Cluster)) %>% 
  mutate(time = as.integer(time)) %>%
  group_by(Cluster, time) %>%
  mutate(m = mean(value)) %>% 
  mutate(sd_min = m-sd(value)) %>% 
  mutate(sd_max = m+sd(value)) %>% 
  mutate(se_min = m-(sd(value)/sqrt(length(value)))) %>% 
  mutate(se_max = m+(sd(value)/sqrt(length(value)))) %>% 
  mutate(quan25 = quantile(value, prob=.25)) %>% 
  mutate(quan75 = quantile(value, prob=.75)) %>%
ggplot() +
  aes(x = time, y = value, group = Cluster, color = Cluster, fill = Cluster, ymin = quan25, ymax = quan75) +
  stat_summary(fun=mean, geom="line", linewidth = 1.5, alpha = 0.7) +
  geom_ribbon(alpha = 0.2) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_x_continuous(breaks = seq(1 , 6, 1), expand = c(0.005,0.005)) +
  labs(x = "Time (hours)", y = "log2 Fold Change", color = "Cluster") +
  theme_bw() +
  theme(legend.position="bottom", text = element_text(size = 12)) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1)


grid.arrange(plota, arrangeGrob(plotb, plotc, heights = c(1,2)), ncol = 2, widths=c(2,1))
```

```{r}
png(width=9, height=8, units="in", res=300, file="data_out/Figure1.png")
grid.arrange(plota, arrangeGrob(plotb, plotc, heights = c(1,2)), ncol = 2, widths=c(2,1))
dev.off()
```

# What are the secondary metabolite biosynthesis genes?

## Secondary metabolites
```{r}
sec_met_go <- as.data.frame(GOTERM) %>% dplyr::select(2:3) %>% filter(str_detect(Term, "terpen|plant volatile|sterol|caroten|saponin|glycoside|phenolic |anthocyani|flav[ao]n|lign[ia]n|coumarin|stilben|tannin|hydroxycinnamate|chalcone|alkaloid|glucosinol|cyanogenic glycoside|phytoalexin|thiosulfinate|pyrrol[i]*|quinoline|pyridine|piperidine|tropane|aporphine|isoquinoline|indolizidine|norlupinane|indole.*purine|purine.*indole|steroid|imidazole")) %>% 
  filter(!str_detect(Term, "catabol|nega|regul|metabol|cholesterol|ergosterol|ergot|lanosterol|steroid hormone secretion|steroid-lactonase activity"))
# note: although present in these GO terms, "corticosteroid" and "ecdysteroid" are not present in our pokeweed annotations
# add column specifying what type it is
sec_met_go$Product <- ifelse(str_detect(sec_met_go$Term, "terpen|plant volatile|sterol|caroten|saponin|glycoside"), "Terpenoids",
                                     ifelse(str_detect(sec_met_go$Term, "phenolic |anthocyani|flav[ao]n|lign[ia]n|coumarin|stilben|tannin|hydroxycinnamate|chalcone"), "Phenolic Compounds",
                                            ifelse(str_detect(sec_met_go$Term, "alkaloid|glucosinol|cyanogenic glycoside|phytoalexin|thiosulfinate|pyrrol[i]*|quinoline|pyridine|piperidine|tropane"), "Nitrogen-sulfur Compounds", NA)))

# add sub_type column
sec_met_go$Sub_type <- ifelse(str_detect(sec_met_go$Term, "terpen"), "Terpenes", ifelse(str_detect(sec_met_go$Term, "plant volatile"), "Plant volatile", ifelse(str_detect(sec_met_go$Term, "sterol"), "Sterols", ifelse(str_detect(sec_met_go$Term, "caroten"), "Carotenoids", ifelse(str_detect(sec_met_go$Term, "saponin"), "Saponins", ifelse(str_detect(sec_met_go$Term, "glycoside"), "Glycosides", ifelse(str_detect(sec_met_go$Term, "anthocyan"), "Anthocyanin", ifelse(str_detect(sec_met_go$Term, "flav[ao]n|anthocyanidin|chalcon"), "Flavonoids", ifelse(str_detect(sec_met_go$Term, "phenolic |hydroxycinnamate|phenylethanoid"), "Phenolics", ifelse(str_detect(sec_met_go$Term, "lign[ia]n"), "Lignin/lignan", ifelse(str_detect(sec_met_go$Term, "coumarin"), "Coumarins", ifelse(str_detect(sec_met_go$Term, "stilben"), "Stilbenes", ifelse(str_detect(sec_met_go$Term, "tannin"), "Tannins",   ifelse(str_detect(sec_met_go$Term, "alkaloid|pyrrol[i]*|quinoline|pyridine|piperidine|tropane"), "Alkaloids", ifelse(str_detect(sec_met_go$Term, "glucosin"), "Glucosinolates", ifelse(str_detect(sec_met_go$Term, "cyanogenic glycoside"), "Cyanogenic glycoside", ifelse(str_detect(sec_met_go$Term, "phytoalexin"), "Phytoalexins", ifelse(str_detect(sec_met_go$Term, "thiosulfinate"), "Thiosulfinate", NA))))))))))))))))))

sec_met_go
```

```{r}
sec_met <- read_tsv("data_out/Supporting_Data_S1.tsv", col_names = c("TranscriptId", "GO")) %>% 
  # select just terms of interest from pokeweed annotations
  filter(str_detect(GO, paste0(sec_met_go$go_id, collapse = "|")))
sec_met
```


```{r}
sec_met <- sec_met %>% 
  # split GO terms
  separate_wider_delim(cols = GO, delim = ",", names_sep = "", too_few = "align_start") %>% 
  # put each GO term on own column
  pivot_longer(contains("GO"), values_to = "go_id") %>% 
  # remove NAs
  filter(!is.na(go_id)) %>% 
  # remove uneeded 'name' column (artifact of separating)
  dplyr::select(!name) %>% 
  # join to GO descriptions
  left_join(sec_met_go, by = "go_id") %>% 
  # remove rows with no relevant GO term
  filter(!is.na(Product)) %>% 
  # remove duplicate sub-type entries for single gene
  group_by(Sub_type, TranscriptId) %>% 
  distinct(Sub_type, .keep_all = T) %>% 
  # add DE data
  left_join(de_genes3) %>% left_join(time_cluster) %>% 
  # put all logFC in one column 
  pivot_longer(contains("logFC"), names_to = "Time", values_to = "logFC") %>% 
  # convert contrast names to just be time
  mutate(Time = as.numeric(str_extract(Time, "[0-9]")))
# if NA then it is not DE
sec_met$DE <- ifelse(is.na(sec_met$Cluster),"No", "Yes")
# replace NAs with 0
sec_met$Cluster[is.na(sec_met$Cluster)] <- 0
sec_met$logFC[is.na(sec_met$logFC)] <- 0
sec_met
```
How many have multiple relevant terms?
```{r}
as.data.frame(table(sec_met$TranscriptId)) %>% arrange(desc(Freq)) %>% 
  # each should have a minimum of 4, for the 4 time points
  mutate(Freq = Freq/4)
```

```{r}
sec_met_pal <- c(colorRampPalette(colors = c("#092a68", "#b0c3d8"))(3), colorRampPalette(colors = c("#5d0baa", "#c2add8"))(5), colorRampPalette(colors = c("#00905d", "#b6ffba"))(5))
as.data.frame(table(sec_met[,c("Product", "Sub_type", "DE")])) %>% 
  # each should have a minimum of 4, for the 4 time points
  mutate(Freq = Freq/4)
sec_met_a <- as.data.frame(table(sec_met[,c("Product", "Sub_type", "DE")])) %>% 
  # each should have a minimum of 4, for the 4 time points
  mutate(Freq = Freq/4) %>%
  mutate(Product_sub = paste0(Product, "-", Sub_type)) %>% 
  filter(Freq > 0) %>% 
  ggplot(aes(x = DE, y = Freq, fill = Product_sub)) +
  geom_col() +
  labs(x = "Differentially expressed", y = "Number of Genes", fill = "Product", title = "A") +
  scale_fill_manual(values = sec_met_pal) +
  theme_minimal()
sec_met_a
```

```{r}
upset_pal <- data.frame(a = sec_met$both, b = ifelse(str_detect(sec_met$both, "Downregulated")&!str_detect(sec_met$both, "Upregulated"), "#d90007",
                                        ifelse(str_detect(sec_met$both, "Upregulated")&!str_detect(sec_met$both, "Downregulated"), "#0069d9",
                                               ifelse(str_detect(sec_met$both, "Downregulated")&str_detect(sec_met$both, "Upregulated"), "#561370", "gray")))) %>% filter(!is.na(a)) %>% distinct(a, .keep_all = T)
upset_pal <- setNames(upset_pal$b, upset_pal$a)
plot(seq_len(length(upset_pal)), 
     rep_len(1, length(upset_pal)),
     col = upset_pal, 
     pch = 16, 
     cex = 6, 
     xaxt = 'n', 
     yaxt = 'n', 
     xlab = '', 
     ylab = '')
```

```{r fig.width = 7, fig.height = 5}
n <- c("Nitrogen-sulfur\nCompounds", "Phenolic\nCompounds", "Terpenoids\n")
names(n) <- c("Nitrogen-sulfur Compounds", "Phenolic Compounds", "Terpenoids")

ns <- sec_met %>% 
  group_by(TranscriptId) %>% 
  filter(Product == "Nitrogen-sulfur Compounds") %>% 
  # exclude non-DE
  filter(Cluster != 0) %>% 
  ggplot() +
  aes(x = Time, y = logFC, fill = TranscriptId, colour = both) +
  geom_line(linewidth = 1.1, alpha = 0.6) +
  scale_color_manual(values = upset_pal) +
  scale_x_continuous(breaks = c(1,2,3,4,5,6)) +
  guides(fill="none", color = "none") +
  labs(color = "Cluster", x = NULL, y = "logFC", title = "B") +
  geom_hline(yintercept = 0) +
  facet_grid(Product~Sub_type, labeller = labeller(Product = n)) + 
  theme_bw() + theme(
    strip.background.y=element_rect(fill=alpha(sec_met_pal[1], 0.4)),
                     strip.text = element_text(size = 9))

pc <- sec_met %>% 
  group_by(TranscriptId) %>% 
  filter(Product == "Phenolic Compounds") %>% 
  # exclude non-DE
  filter(Cluster != 0) %>% 
  ggplot() +
  aes(x = Time, y = logFC, fill = TranscriptId, colour = both) +
  geom_line(linewidth = 1.1, alpha = 0.6) +
  scale_color_manual(values = upset_pal) +
  scale_x_continuous(breaks = c(1,2,3,4,5,6)) +
  guides(fill="none", color = "none") +
  labs(color = "Cluster", x = NULL, y = "logFC") +
  geom_hline(yintercept = 0) +
  facet_grid(Product~Sub_type, labeller = labeller(Product = n)) + 
  theme_bw() + theme(
    strip.background.y=element_rect(fill=alpha(sec_met_pal[4], 0.4)),
                     strip.text = element_text(size = 10))

tr <- sec_met %>% 
  group_by(TranscriptId) %>% 
  filter(Product == "Terpenoids") %>% 
  # exclude non-DE
  filter(Cluster != 0) %>% 
  ggplot() +
  aes(x = Time, y = logFC, fill = TranscriptId, colour = both) +
  geom_line(linewidth = 1.1, alpha = 0.6) +
  scale_color_manual(values = upset_pal) +
  scale_x_continuous(breaks = c(1,2,3,4,5,6)) +
  guides(fill="none", color = "none") +
  labs(color = "Cluster", x = "Time (Hours)", y = "logFC") +
  geom_hline(yintercept = 0) +
  facet_grid(Product~Sub_type, labeller = labeller(Product = n)) + 
  theme_bw() + theme(
    strip.background.y=element_rect(fill=alpha(sec_met_pal[9], 0.4)),
                     strip.text = element_text(size = 10))

sec_met_b <- grid.arrange(ns, pc, tr, ncol = 1)
sec_met_b
```

```{r fig.width = 10, fig.height = 9}
png(width=8, height=6, units="in", res=300, filename = "Figure4.png")
grid.arrange(sec_met_a +
               theme(legend.position = "bottom",legend.title=element_blank())+
               guides(fill=guide_legend(ncol=1)), 
             sec_met_b, nrow = 1, widths = c(2,3))
dev.off()
```

# Heatmap only Arabidopsis genes of interest

## Get genes

```{r}
arab_genes <- read.fasta("JA_pathway/Araport11_pep_20220914_representative_gene_model.fa", as.string = T, seqtype = "AA", whole.header = T)
arab_genes <- data.frame(TranscriptId = names(arab_genes), Seq = paste(arab_genes))
arab_genes
```

```{r}
arab_genes <- arab_genes %>% 
  # filter out starting sequence
  mutate(TranscriptId = str_extract(TranscriptId, "Symbols: [A-Za-z0-9, -.]*")) %>% 
  # select only genes of interest
  # looking for ones starting with whitespace, then gene name, then either a comma or whitespace followed by the end of the line
  filter(str_detect(TranscriptId, paste0("\\s", read_csv("JA_pathway/genes.csv")$Gene, "|\\s", read_csv("JA_pathway/genes.csv")$Gene2, collapse = "|"))) %>% 
  # get rid of "Symbols: "
  mutate(TranscriptId = str_remove(TranscriptId, "Symbols: ")) %>% 
  # extract just gene ID
  mutate(TranscriptId = str_extract(TranscriptId, paste0(read_csv("JA_pathway/genes.csv")$Gene, "|", read_csv("JA_pathway/genes.csv")$Gene2, collapse = "|"))) %>% 
  # remove extra commas
  mutate(TranscriptId = str_remove(TranscriptId, ","))
arab_genes
```

```{r}
write.fasta(strsplit(arab_genes$Seq,""), arab_genes$TranscriptId, file = "JA_pathway/arab_proteins.fasta")
```

Command line for blast:

```
cd /home/kyrad/JA2022_RNAseq/arab_blast
ln -s /home/kyrad/new_assembly/step7_genome_annotation/combined/annotations_combined_Nov2022_pep.fa

# blast the known rips (happen to be amino acid sequences because that is what I have)
blastp -query arab_proteins.fasta -subject annotations_combined_Nov2022_pep.fa -evalue 0.01 -outfmt 10 -out blast_out_arab.csv
```

```{r}
de_JA <- read_csv("JA_pathway/blast_out_arab.csv", col_names = c("query", "TranscriptId", "percent_identical", "alignment_length", "number_mismatches", "gapopenings", "qstart", "qend", "sstart", "send", "evalue", "bitscore")) %>% 
  # tidy TranscriptId
  mutate(TranscriptId = str_remove(TranscriptId, "_[0-9]")) %>% 
  # merge with logFC values
  left_join(., de_genes, by = "TranscriptId") %>% 
  # keep only those with high percent ID
  filter(percent_identical > 50) %>% 
  # keep only those with bit score > 40
  filter(bitscore > 40) %>% 
  # remove duplicate pokeweed genes (keep only the transcript IDs with the highest bit score)
  arrange(desc(bitscore)) %>% 
  arrange(desc(percent_identical)) %>% 
  distinct(TranscriptId, .keep_all = T) %>% 
  # concatenate arabidopsis name and pokeweed name
  mutate(TID = paste0(query,"-",TranscriptId)) %>% 
  column_to_rownames(var = "TID") %>% 
  mutate_if(is.numeric, ~replace_na(., 0)) %>% 
  group_by(query) %>% 
  slice_max(percent_identical, n=10) %>% 
  arrange(query) %>% 
  mutate(TID = paste0(query,"-",TranscriptId)) %>% 
  column_to_rownames(var = "TID")

de_JA
```

```{r}
write_csv(rownames_to_column(de_JA[,1:12], var = "TranscriptID"), "data_out/Supporting_Data_S4.csv")
```

## Heatmap

```{r}
colors <- c(RColorBrewer::brewer.pal(7, "YlOrBr")[6:2], "white", mako(18, direction = -1))
```

```{r, fig.width = 4.1, fig.height = 15}
de_JA %>% 
  group_by(TranscriptId) %>% 
  filter(sum(logFC_JAvsEt1,logFC_JAvsEt1,logFC_JAvsEt1,logFC_JAvsEt1)!=0) %>% 
  mutate(TID = paste0(query,"-",TranscriptId)) %>% 
  column_to_rownames(var = "TID") %>% 
  dplyr::select(13:16) %>% 
  pheatmap(., legend_breaks = c(-4:20), fontsize = 12, color = colors, cutree_rows = 1, show_rownames = T, angle_col = "90", cluster_cols = F, cluster_rows=F, main = "logFC of JA-related genes")
```

# Upset plot all time points
```{r fig.width = 8, fig.height = 5}
de_genes3 <- pairwise_de %>% filter(!(FDR_JAvsEt1 >= fdr & FDR_JAvsEt2 >= fdr & FDR_JAvsEt4 >= fdr & FDR_JAvsEt6 >= fdr))
x = list("1h.Upregulated" = de_genes3$TranscriptId[de_genes3$logFC_JAvsEt1 > lfc & de_genes3$FDR_JAvsEt1 < fdr], 
         "2h.Upregulated" = de_genes3$TranscriptId[de_genes3$logFC_JAvsEt2 > lfc & de_genes3$FDR_JAvsEt2 < fdr], 
         "4h.Upregulated" = de_genes3$TranscriptId[de_genes3$logFC_JAvsEt4 > lfc & de_genes3$FDR_JAvsEt4 < fdr], 
         "6h.Upregulated" = de_genes3$TranscriptId[de_genes3$logFC_JAvsEt6 > lfc & de_genes3$FDR_JAvsEt6 < fdr], 
         "1h.Not_DE" = de_genes3$TranscriptId[lfc & de_genes3$FDR_JAvsEt1 >= fdr],
         "2h.Not_DE" = de_genes3$TranscriptId[lfc & de_genes3$FDR_JAvsEt2 >= fdr],
         "4h.Not_DE" = de_genes3$TranscriptId[lfc & de_genes3$FDR_JAvsEt4 >= fdr],
         "6h.Not_DE" = de_genes3$TranscriptId[lfc & de_genes3$FDR_JAvsEt6 >= fdr],
         "1h.Downregulated" = de_genes3$TranscriptId[de_genes3$logFC_JAvsEt1 < -lfc & de_genes3$FDR_JAvsEt1 < fdr], 
         "2h.Downregulated" = de_genes3$TranscriptId[de_genes3$logFC_JAvsEt2 < -lfc & de_genes3$FDR_JAvsEt2 < fdr],
         "4h.Downregulated" = de_genes3$TranscriptId[de_genes3$logFC_JAvsEt4 < -lfc & de_genes3$FDR_JAvsEt4 < fdr], 
         "6h.Downregulated" = de_genes3$TranscriptId[de_genes3$logFC_JAvsEt6 < -lfc & de_genes3$FDR_JAvsEt6 < fdr])
```

```{r}
paste0("Number of genes: ",nrow(de_genes3))
```
Output:
```
"Number of genes: 5133"
```

Note: this is Figure 2 (the 'groups' were added to the bottom in powerpoint)
```{r fig.width = 8, fig.height = 7.8}
upset(fromList(x), 
      sets = c("1h.Upregulated", "2h.Upregulated", "4h.Upregulated", "6h.Upregulated", "1h.Not_DE", "2h.Not_DE", "4h.Not_DE", "6h.Not_DE", "1h.Downregulated", "2h.Downregulated", "4h.Downregulated", "6h.Downregulated"), 
      order.by = "freq", 
      keep.order = T, 
      nintersects = 20,
      nsets = 60, 
      line.size = 1, 
      empty.intersections = "on", 
      text.scale = 1.6,
      sets.bar.color = c(rep("#054fb5",4), rep("#4d4d4d", 4), rep("#912301", 4,)),
      main.bar.color = c("#d90007", "#0069d9", "#d90007", "#d90007", "#0069d9", "#0069d9", "#d90007", "#0069d9", "#0069d9", "#d90007", "#0069d9", "#d90007", "#d90007", "#0069d9", "#0069d9", "#d90007", "#0069d9", "#d90007", "#561370", "#0069d9", "#0069d9", "#d90007", "#d90007", "#561370", "#d90007", "#d90007", "#0069d9", "#0069d9", "#d90007", "#561370", "#561370", "#561370", "#d90007", "#561370", "#561370", "#561370", "#d90007", "#561370", "#561370", "#561370", rep("gray", 4055)))
```

# GO analysis of de subsets


```{r}
de_genes3 <- pairwise_de %>% 
  filter(!(FDR_JAvsEt1 >= fdr & FDR_JAvsEt2 >= fdr & FDR_JAvsEt4 >= fdr & FDR_JAvsEt6 >= fdr)) %>% 
  mutate(JAvsEt1 = ifelse(logFC_JAvsEt1 > lfc & FDR_JAvsEt1 < fdr, "Upregulated", 
                          ifelse(logFC_JAvsEt1 < -lfc & FDR_JAvsEt1 < fdr, "Downregulated", "Not_DE"))) %>% 
  mutate(JAvsEt2 = ifelse(logFC_JAvsEt2 > lfc & FDR_JAvsEt2 < fdr, "Upregulated", 
                          ifelse(logFC_JAvsEt2 < -lfc & FDR_JAvsEt2 < fdr, "Downregulated", "Not_DE"))) %>% 
  mutate(JAvsEt4 = ifelse(logFC_JAvsEt4 > lfc & FDR_JAvsEt4 < fdr, "Upregulated", 
                          ifelse(logFC_JAvsEt4 < -lfc & FDR_JAvsEt4 < fdr, "Downregulated", "Not_DE"))) %>%
  mutate(JAvsEt6 = ifelse(logFC_JAvsEt6 > lfc & FDR_JAvsEt6 < fdr, "Upregulated", 
                          ifelse(logFC_JAvsEt6 < -lfc & FDR_JAvsEt6 < fdr, "Downregulated", "Not_DE"))) %>%
  unite("both", c(JAvsEt1, JAvsEt2, JAvsEt4, JAvsEt6), sep = "-")

de_genes3
```

# GO analysis per group
```{r}
geneID2GO <- readMappings(file = "data_out/Supporting_Data_S1.tsv")
str(head(geneID2GO))
geneNames <- names(geneID2GO)
```

```{r}
n <- unique(de_genes3$both)
sig_go_terms_g3 <- data.frame(GO.ID = character(), Term = character(), Annotated = integer(), Significant = integer(), Expected = double(), "Rank in classic" = integer(), classic = character(), elim = character(), weight = character(), Group = character())

for (i in 1:length(n)) {
  myInterestingGenes <- de_genes3$TranscriptId[de_genes3$both == n[i]]
  geneList <- factor(as.integer(geneNames %in% myInterestingGenes))
  names(geneList) <- geneNames
  
  GOdata <- new("topGOdata", ontology = "BP", allGenes = geneList, annot = annFUN.gene2GO, gene2GO = geneID2GO, nodeSize = 8)
  
  resultFis <- runTest(GOdata, algorithm = "classic", statistic = "fisher")
  resultFisher <- runTest(GOdata, algorithm = "elim", statistic = "fisher")
  resultWeight <- runTest(GOdata, algorithm = "weight", statistic = "fisher")
  
  allRes <- GenTable(GOdata, classic = resultFis, elim = resultFisher, weight = resultWeight, orderBy = "weight", ranksOf = "classic", topNodes = 25)
  allRes$Group <- n[i]
  
  sig_go_terms_g3 <- rbind(sig_go_terms_g3, allRes)
}
```


```{r}
sig_go_terms3 <- sig_go_terms_g3 %>% 
  mutate(classic = as.numeric(classic)) %>% 
  mutate(elim = as.numeric(elim)) %>% 
  mutate(weight = as.numeric(weight)) %>% 
  filter(elim < 0.01) %>% 
  filter(weight < 0.01) %>% 
  filter(classic < 0.01)

sig_go_terms3
```

```{r}
write_csv(sig_go_terms3, "data_out/Supporting_Data_S3.csv")
```

# RIPs

## How many genes identified as a RIP?

The RIP names we know:
```{r}
rip_names <- read_csv("data_in/blast_out_rips.csv") %>% dplyr::select(1:2) %>% rename("qaccver" = "Common", "saccver" = "TranscriptId")
rip_names
```

```{r}
rips <- interpro %>% filter(str_detect(InterPro_annotation_description, "[Rr]ibosome[- ][Ii]nacti")) %>% filter(Analysis == "Pfam")
rips <- data.frame(TranscriptId = c(rips$Protein_accession, rip_names$TranscriptId)) %>% 
  distinct(TranscriptId) %>% 
  full_join(rip_names, by = "TranscriptId") %>% 
  mutate(Common = replace_na(Common, "Uncharacterized"))
rips %>% arrange(Common)
```


```{r}
rips <- left_join(rips, de_genes3, by = "TranscriptId") %>% left_join(time_cluster, by = "TranscriptId") %>% 
  pivot_longer(contains("logFC"), names_to = "Time", values_to = "logFC") %>% 
  mutate(Time = as.numeric(str_extract(Time, "[0-9]")))
rips %>% dplyr::select(c(1:2,11:14))
```

```{r}
rip_pal <- setNames(turbo(7)[1:6], sort(unique(rips$Common[!is.na(rips$Cluster)])))
plot(seq_len(length(rip_pal)), 
     rep_len(1, length(rip_pal)),
     col = rip_pal, 
     pch = 16, 
     cex = 6, 
     xaxt = 'n', 
     yaxt = 'n', 
     xlab = '', 
     ylab = '')
```

```{r}
png(width=5, height=4, res=300, units="in", file = "data_out/Figure3.png")
rips %>% 
  group_by(TranscriptId) %>% 
  filter(!is.na(Cluster)) %>% 
  ggplot() +
  aes(x = Time, y = logFC, shape = TranscriptId, color = Common) +
  geom_line(linewidth = 1.5, alpha = 0.6) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0) +
  guides(fill="none") +
  labs(color = "Gene", x = "Time (Hours)", y = "logFC") +
  scale_x_continuous(breaks = seq(1 , 6, 1), expand = c(0,0)) +
  scale_color_manual(values = rip_pal) +
  scale_shape_manual(values = c(4,21:25)) +
  theme_bw()
dev.off()
```

# Compare assemblies 2019 vs 2022
```{r}
path_19 <- "data_in/maker_annotation.gff3"
gff3_19 <- read_tsv(path_19, col_names = NULL, comment = "#")

gff3_22 <- gff3
```

```{r}
g_19 <- gff3_19 %>% 
  mutate(ID = str_extract(X9, "[A-Z]+_[0-9]+-[A-Z]{2}")) %>% 
  # only interested in primary transcript
  filter(str_detect(ID, "RA")) %>% 
  dplyr::select(c(3:5,10)) %>% 
  # remove UTRs
  filter(!str_detect(X3, "UTR")) %>% 
  # remove genes
  filter(X3 != "gene") %>% 
  # get length
  mutate(Length = abs(X4-X5)) %>% 
  dplyr::select(!c(X4, X5))
# remove IDs associated with genes
g_19
```

```{r}
s_19 <- g_19 %>% pivot_wider(names_from = X3, values_from = Length) %>% 
  group_by(ID) %>% 
  # convert mRNA from list to number
  mutate(mRNA = as.numeric(mRNA)) %>% 
  # get number of elements per transcript
  mutate(n_CDS = lengths(CDS)) %>% 
  mutate(n_exon = lengths(exon)) %>% 
  # get total element length
  mutate(len_CDS = sum(unlist(CDS))) %>% 
  mutate(len_exon = sum(unlist(exon))) %>% 
  mutate(len_intron = mRNA-len_exon) %>% 
  # get mean element length
  mutate(mean_CDS = mean(unlist(CDS))) %>% 
  mutate(mean_exon = mean(unlist(exon))) %>% 
  # remove uneeded column
  dplyr::select(!c(exon, CDS))
s_19
```

```{r}
sum_19 <- s_19 %>% 
  ungroup() %>% 
  dplyr::select(!ID) %>% 
  pivot_longer(1:8, names_to = "Feature") %>% 
  group_by(Feature) %>% 
  # remove NAs
  filter(!is.na(value)) %>% 
  # generate summary stats
  summarize(Mean = mean(value), Minimum = min(value), Maximum = max(value), Median = median(value)) %>% 
  mutate(Assembly = "2019")
sum_19
```

```{r}
g_22 <- gff3_22 %>% 
  mutate(ID = str_extract(X9, "anno[0-9]+\\.[a-z]+[0-9]+\\.[a-z]+[0-9]+")) %>% 
  dplyr::select(c(3:5,10)) %>% 
  # get length
  mutate(Length = abs(X4-X5)) %>% 
  dplyr::select(!c(X4, X5))
# rename transcript to mRNA
g_22$X3[g_22$X3 == "transcript"] <- "mRNA"
g_22
```

```{r}
s_22 <- g_22 %>% pivot_wider(names_from = X3, values_from = Length) %>% 
  group_by(ID) %>% 
  # convert mRNA from list to number
  mutate(mRNA = as.numeric(mRNA)) %>% 
  # get number of elements per mRNA
  mutate(n_CDS = lengths(CDS)) %>% 
  mutate(n_exon = lengths(exon)) %>% 
  # get total element length
  mutate(len_CDS = sum(unlist(CDS))) %>% 
  mutate(len_exon = sum(unlist(exon))) %>% 
  mutate(len_intron = mRNA-len_exon) %>% 
  # get mean element length
  mutate(mean_CDS = mean(unlist(CDS))) %>% 
  mutate(mean_exon = mean(unlist(exon))) %>% 
  # remove uneeded column
  dplyr::select(!c(exon, CDS))
s_22
```

```{r}
sum_22 <- s_22 %>% 
  ungroup() %>% 
  dplyr::select(!ID) %>% 
  pivot_longer(1:8, names_to = "Feature") %>% 
  group_by(Feature) %>% 
  # remove NAs
  filter(!is.na(value)) %>% 
  # generate summary stats
  summarize(Mean = mean(value), Minimum = min(value), Maximum = max(value), Median = median(value)) %>% 
  mutate(Assembly = "2022")
sum_22
```

```{r}
comb <- rbind(sum_19, sum_22) %>% 
  mutate(Feature = str_replace(Feature, "len_CDS", "Total length of CDS")) %>% 
  mutate(Feature = str_replace(Feature, "len_exon", "Total length of exons")) %>% 
  mutate(Feature = str_replace(Feature, "len_intron", "Total length of introns")) %>%
  mutate(Feature = str_replace(Feature, "^n_CDS", "Number of CDS per gene")) %>% 
  mutate(Feature = str_replace(Feature, "^n_exon", "Number of exons per gene")) %>% 
  mutate(Feature = str_replace(Feature, "mean_CDS", "Mean length of CDS")) %>% 
  mutate(Feature = str_replace(Feature, "mean_exon", "Mean length of exons")) %>% 
  mutate(Feature = str_replace(Feature, "mRNA", "Length of mRNA"))
comb
```

```{r}
s1 <- comb %>%
 filter(Feature %in% c("Length of mRNA", "Total length of CDS", "Total length of exons", "Total length of introns"
)) %>%
 ggplot() +
  aes(x = Feature, fill = Assembly, weight = Mean) +
  geom_bar(position = "dodge") +
  labs(y = "Mean Length (bp)", x = "") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal()
```

```{r}
s2 <- rbind(mutate(s_19, Assembly = "2019"), mutate(s_22, Assembly = "2022")) %>% 
  ungroup() %>% 
  dplyr::select(!ID) %>% 
  pivot_longer(1:8) %>% 
  mutate(name = str_replace(name, "len_CDS", "Total length of CDS")) %>% 
  mutate(name = str_replace(name, "len_exon", "Total length of exons")) %>% 
  mutate(name = str_replace(name, "len_intron", "Total length of introns")) %>%
  mutate(name = str_replace(name, "^n_CDS", "Number of CDS")) %>% 
  mutate(name = str_replace(name, "^n_exon", "Number of exons")) %>% 
  mutate(name = str_replace(name, "mean_CDS", "Mean length of CDS")) %>% 
  mutate(name = str_replace(name, "mean_exon", "Mean length of exons")) %>% 
  mutate(name = str_replace(name, "mRNA", "Length of mRNA")) %>% 
 filter(name %in% c("Length of mRNA", "Total length of CDS", "Total length of exons", "Total length of introns"
)) %>%
 ggplot() +
  aes(x = name, y = value, fill = Assembly) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set1", direction = 1) +
  scale_y_continuous(trans = "log10") +
  labs(x = "Feature", y = "Log10 Feature Length (bp)") +
  theme_minimal()
```

```{r}
png(width=7, height=5, units="in", res=300, file="data_out/Supporting_Figure_S2.png")
grid.arrange(s1+labs(title="A"), s2+labs(title="B"), ncol=1)
dev.off()
```
